---
title: "Visualization"
author: "Janine Melsen"
date: "08/04/2020"
output: html_document
---

In this markdown document, we will demonstrate how you could visualize the different methods for clustering and dimensionality reduction, as stored in the data frame (the output from the clustering_dimensionalityreduction_pseudotime script/markdown)

As input we will use the deposited 'df.csv' file, which contains the expression values of 275856 CD4 T cells, HSNE-based clustres, flowsom clusters, phenograph clusters, umap coordinates, diffusion map coordinates and pseudotime values.

## Load packages
```{r library, warning=FALSE, message=FALSE, error=FALSE}
library(dplyr)
library(ggplot2)
library(scales)
library(reshape2)
library(RColorBrewer)
library(ggrepel)
```


## Load dataframe

```{r load data, echo=FALSE}
df<- read.csv('//vf-d3-home/d3home$/jemelsen/MyDocs/GitHub/Single-cell-analysis-flow-cytometry/df.csv', row.names=1)

head(df)
```


## Visualization
### Visualize parameters on diffusion map

Define a function for visualization of the diffusion map
```{r visualization function}
viz.dm <- function(dat,dr, param.name,limits=NULL){
  ColVal <- dat[,param.name]
  if(is.null(limits)){
    Lim <- quantile(ColVal,probs=seq(0,1,0.01))[c(2,100)]
    p <- ggplot(dat, aes(x = DC1, y =DC2)) +geom_point(aes(color = ColVal), size=0.1)+theme_classic()+scale_color_distiller(name=param.name, palette = "RdYlBu", limits=Lim, oob=squish)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+ggtitle(param.name)
  } else {
    p <- ggplot(dat, aes(x = DC1, y = DC2)) +geom_point(aes(color = ColVal), size=0.1)+theme_classic()+scale_color_distiller(name=param.name, palette = "RdYlBu", limits=c(limits[1],limits[2]), oob=squish)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+ggtitle(param.name)
  }
  p
}
```

Define the parameters for which you would like generate plots
First check the order of columns
```{r parameter definition}
colnames(df)
```

Provide the column numbers
```{r define parameters}
parameters = colnames(df)[c(7:9, 11, 13:16,18,19)]
```

Visualize the parameters on the diffusion map
```{r vizualize dm, fig.height=3, fig.width=4.5}
for (i in 1:length(parameters)){
  p=viz.dm(dat=df, param.name = parameters[i])
  print(p)
}
```

If you are not satisfied with the color scale, adjust the limits argument
```{r use limits,fig.height=3, fig.width=4.5}
viz.dm(dat=df,param.name='CD8', limits=c(-0.36,4.32))
viz.dm(dat=df, param.name='CD103', limits=c(-0.38, 3.16))
```

### visualize clusters on diffusion map

```{r clusters diffusion map, message=FALSE, fig.height=3, fig.width=4}
label_HSNE_dm <- df%>%group_by(clusters_HSNE)%>%select(DC1, DC2)%>%summarize_all(mean)
label_flowsom_dm <- df%>%group_by(clusters_flowsom)%>%select(DC1, DC2)%>%summarize_all(mean)
label_pheno_dm <- df%>%group_by(clusters_phenograph)%>%select(DC1, DC2)%>%summarize_all(mean)

ggplot(df, aes(x=DC1, y=DC2, color=as.factor(clusters_HSNE)))+geom_point(size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+geom_label_repel(aes(label=clusters_HSNE), data=label_HSNE_dm)+guides(colour=FALSE)+ggtitle('clusters HSNE')

ggplot(df, aes(x=DC1, y=DC2, color=as.factor(clusters_flowsom)))+geom_point(size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+geom_label_repel(aes(label=clusters_flowsom), data=label_flowsom_dm)+guides(colour=FALSE)+ggtitle('clusters flowsom')

ggplot(df, aes(x=DC1, y=DC2, color=as.factor(clusters_phenograph)))+geom_point(size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+geom_label_repel(aes(label=clusters_phenograph), data=label_pheno_dm)+guides(colour=FALSE)+ggtitle('clusters phenograph')

```

### visualize parameters on umap

Define a function for visualization of the umap
```{r parameters umap, echo=FALSE}
viz.umap <- function(dat,param.name,limits=NULL){
  ColVal <- dat[,param.name]
  if(is.null(limits)){
    Lim <- quantile(ColVal,probs=seq(0,1,0.01))[c(2,100)]
    p <- ggplot(dat, aes(x = umap_1, y =umap_2)) +geom_point(aes(color = ColVal), size=0.1)+theme_classic()+scale_color_distiller(name=param.name, palette = "RdYlBu", limits=Lim, oob=squish)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+ggtitle(param.name)
  } else {
    p <- ggplot(dat, aes(x = umap_1, y = umap_2)) +geom_point(aes(color = ColVal), size=0.1)+theme_classic()+scale_color_distiller(name=param.name, palette = "RdYlBu", limits=c(limits[1],limits[2]), oob=squish)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+ggtitle(param.name)
  }
  p
}

```

Visualize the parameters on the umap
```{r vizualize umap, fig.height=3, fig.width=4}
for (i in 1:length(parameters)){
  p=viz.umap(dat=df, param.name = parameters[i])
  print(p)
}
```

If you are not satisfied with the color scale, adjust the limits argument
```{r use limits umap, fig.height=3, fig.width=4}
viz.umap(dat=df,param.name='CD8', limits=c(-0.36,4.32))
viz.umap(dat=df, param.name='CD103', limits=c(-0.38, 3.16))
```


### visualize clusters on umap

```{r warning=FALSE, message=FALSE, fig.height=3, fig.width=4}
label_HSNE_umap <- df%>%group_by(clusters_HSNE)%>%select(umap_1, umap_2)%>%summarize_all(mean)
label_flowsom_umap <- df%>%group_by(clusters_flowsom)%>%select(umap_1, umap_2)%>%summarize_all(mean)
label_pheno_umap <- df%>%group_by(clusters_phenograph)%>%select(umap_1, umap_2)%>%summarize_all(mean)

ggplot(df, aes(x=umap_1, y=umap_2, color=as.factor(clusters_HSNE)))+geom_point(size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+geom_label_repel(aes(label=clusters_HSNE), data=label_HSNE_umap)+guides(colour=FALSE)+ggtitle('clusters HSNE')

ggplot(df, aes(x=umap_1, y=umap_2, color=as.factor(clusters_flowsom)))+geom_point(size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+geom_label_repel(aes(label=clusters_flowsom), data=label_flowsom_umap)+guides(colour=FALSE)+ggtitle('clusters flowsom')

ggplot(df, aes(x=umap_1, y=umap_2, color=as.factor(clusters_phenograph)))+geom_point(size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+geom_label_repel(aes(label=clusters_phenograph), data=label_pheno_umap)+guides(colour=FALSE)+ggtitle('clusters phenograph')
```

### Frequencies

Fill in for which clusters you would like to calculate the frequencies (either clusters_HSNE, clusters_flowsom, or clusters_phenograph)
```{r counts df}
counts <- as.data.frame.matrix(table(df$sampleID,df$clusters_HSNE))
```

Calculate the percentage of a certain sample present in a certain cluster
```{r counts}
counts_percofsample = counts/rowSums(counts)*100
counts_percofsample$total <- rowSums(counts)
counts_percofsample$sampleID <- row.names(counts_percofsample)
counts_percofsample <- melt(counts_percofsample,  id.vars=c('sampleID', 'total'), variable.name='cluster', value.name='frequency')
```

Plot the frequencies in a boxplot (frequency = % of sample in cluster)
```{r boxplot}
ggplot(counts_percofsample, aes(x=reorder(cluster, frequency, FUN=median), y=frequency))+ geom_boxplot(position='dodge2', outlier.shape=NA)+geom_jitter(aes(colour=sampleID), size=2)+xlab('cluster')
```


Plot the frequencies in bargraphs (frequency = % of sample in cluster)
```{r, echo=FALSE}
ggplot(counts_percofsample, aes(x=cluster, y=frequency, fill=cluster))+geom_bar(stat='identity')+facet_wrap(~sampleID)+theme(axis.text.x=element_text(angle=30, vjust=.9, hjust=1))
```

### Visualize pseudotime

######If you would like to use other clusters, adjust 'merged_HSNE' to clusters_flowsom', for instance
######If there are more or less lineages identified adjust the number of curve columns
```{r pseudotime}
pseudotimevalues <- df%>%select(c(DC1, DC2,clusterID=merged_HSNE, curve1, curve2, curve3))
```

Plot the lineages: cells are colored by pseudotime
```{r plot lineages}
#reshape table 
pseudotimevalues <- melt(pseudotimevalues, id.vars=c('clusterID', 'DC1', 'DC2'), variable.name='lineage', value.name = 'pseudotime')

#rename curve to lineage
pseudotimevalues$lineage <- gsub('curve','lineage', pseudotimevalues$lineage)

#exclude cells with NA pseudotimevalues (those cells are not present in all lineages and have NA values for the lineages in which they are absent)
pseudotimevaluesexclNA <- pseudotimevalues%>%filter(pseudotime!='NA')

#generate colorpalette
colors <- colorRampPalette(rev(brewer.pal(11, 'Spectral'))[-6])(100)

#ggplot of each lineage colored by either pseudotime
ggplot(pseudotimevaluesexclNA%>%arrange(pseudotime), aes(x=DC1, y=DC2))+geom_point(aes(color=pseudotime),size=0.1, alpha=0.3) +facet_wrap(~lineage)+scale_color_gradientn(colours=colors)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())
```

Plot the lineages: cells are colored by clusterID
```{r, fig.height=3, fig.width=10}
ggplot(pseudotimevaluesexclNA%>%arrange(pseudotime), aes(x=DC1, y=DC2))+geom_point(aes(color=clusterID),size =0.1)+facet_wrap(~lineage)+theme_bw()+ guides(colour = guide_legend(override.aes = list(size=5)))
```


Jitterplot per lineage: ells are colored by clusterID
```{r jitterplot, fig.height=3, fig.width=6}
ggplot(pseudotimevaluesexclNA%>%arrange(pseudotime), aes(x=pseudotime, y=lineage)) + geom_jitter(aes(color=clusterID), size=0.1)+theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor=element_blank())+ guides(colour = guide_legend(override.aes = list(size=5)))
```

